---
title: "20230502_uafR_test_Liz"
author: "Olivia Trase"
date: "2023-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

#### Testing the uafR Package on maize volatile data ####

Elizabeth Davidson-Lowe ran an experiment using maize, western corn rootworm, and nematodes. We analyzed the aboveground volatiles using GC-MS. 


## Loading in libraries and data
```{r}
## install packages if needed
# install.packages("devtools")
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install("fmcsR")
# devtools::install_github("castratton/uafR")

## load libraries
library(fmcsR)
library(ChemmineR)
library(uafR)
library(vegan)
library(ggplot2)
library(tidyr)
library(dplyr)
library(reshape2)
library(gt)

```

```{r}
## load in data
voc <-read.csv("20230411_MassHunter_Hits.csv")
head(voc)
```
This data was exported from MassHunter using default deconvolution settings and the chemecol, Adams, and NIST17 spectral libraries.

```{r}
Nate <- read.csv("AMFWCREPN.csv")
head(Nate)
## clean data
voc$Compound.Name <- gsub("?","",voc$Compound.Name, fixed = TRUE)

```
This data was compiled and peaks were manually identified and peak areas calculated by Nate McCartney.

## Run spreadOut, mzExacto, and standardifyIt
Can load in Target_Compounds.csv if spreadOut takes too long
```{r}
## uafR test of SpreadOut and mzExacto
## load sample library
library_dat <- library_data 

## spread out data, interface with pubchem
voc_spread <- spreadOut(voc) # takes like 6-7 hours

## name compounds of interest
## these compounds were originally identified by Nate
query_chemicals <- c("(E)-beta-ocimene", "(Z)-3-hexenyl acetate", "1-dodecanol", "2,6-di-tert-butyl-1","4-benzoquinone",
                     "2-heptanone", "3-heptanone", "alpha-pinene", "benzaldehyde", "beta-pinene", "camphene", "camphor", 
                     "Cumene", "delta-3-carene", "hexanal 2-ethyl-", "limonene", "linalool", "methyl salicylate", "naphthalene", 
                     "n-decanal", "n-heptanal", "n-hexanal", "n-nonanal", "n-octanal", "n-octane", "nonyl acetate",
                     "n-tridecan-1-ol","p-cymene" , "pentanone<4-hydroxy-4-methyl-2->", "sabinene", "tetradecanal")
print(query_chemicals)
## mzExacto: compile queried chemicals into matrix where rows are the compounds and columns are samples
voc_crunch <- mzExacto(data_in = voc_spread, chemicals = query_chemicals)
# write.csv(voc_crunch, "Target_Compounds.csv") # save a csv midway through

## Normalize uafR compounds
# voc_crunch<-read.csv("Target_Compounds.csv") # load saved csv
voc_normalized <- standardifyIt(data_in = voc_crunch, standard_type = "Internal", 
                                standard_used = "nonyl acetate", 
                                IS_ng = 200, IS_uL = 5) # added 5 microliters of 40ng/ul nonyl-acetate standard

```
These are the compounds that Nate identified 

## Comparison between uafR and Manual identification
```{r}
## clean up Nate's data
Nate$Analyte <- trimws(Nate$Analyte, "r")

## pivot the df to get rows as compounds, columns as samples, values as quantity in ng
Nate_pivot <- pivot_wider(data = Nate, 
                          names_from = Filename, 
                          values_from = Quantity, 
                          id_cols = Analyte,
                          names_prefix = "")

## get the retention times for each compound
Nate_pivotRT <- pivot_wider(data = Nate, 
                          names_from = Filename, 
                          values_from = RT, 
                          id_cols = Analyte,
                          names_prefix = "")

## remove the controls
Nate_pivotRT <- Nate_pivotRT[c(1,10:68)]
Nate_pivotRT[Nate_pivotRT == 0] <- NA # 0 retention time is really NA

## get the average retention time
RTmeans <- rowMeans(Nate_pivotRT[c(2:60)],na.rm=TRUE)
apply(Nate_pivotRT[c(2:60)], 1, sd,na.rm=TRUE)   # check that standard deviation is not huge, looks good

## Add RT column for rounded values
Nate_pivot$RT <- round(RTmeans,2)

## make column names between two dataframes the same
colnames(Nate_pivot)<-gsub("-",".",colnames(Nate_pivot))
names(Nate_pivot)[names(Nate_pivot) == 'Analyte'] <- 'Compound'

# ## get list of sample names
# Nate_samples <- colnames(Nate_pivot[c(10:68)])
# voc_samples <- colnames(voc_normalized[c(6:65)])

## call columns of interest
Nate_new <- Nate_pivot[c(1,10:69)] 
VOC_new <- voc_normalized[c(2,4,6:64)] # remove T2-5

## Get compounds that do not match between two datasets
print(c(setdiff(VOC_new$Compound,Nate_new$Compound),setdiff(Nate_new$Compound,VOC_new$Compound)))

## Rename compounds that are the same
Nate_new$Compound<-gsub("cumene","Cumene",Nate_new$Compound)
Nate_new$Compound<-gsub("napthalene","naphthalene",Nate_new$Compound)

## Remove rest
remove_list <- c(setdiff(VOC_new$Compound,Nate_new$Compound),setdiff(Nate_new$Compound,VOC_new$Compound))
print(remove_list)
Nate_new1 <- Nate_new[ !grepl(paste(remove_list, collapse="|"), Nate_new$Compound),]

# Check that datasets match
setdiff(colnames(Nate_new1),colnames(VOC_new))
setdiff(VOC_new$Compound,Nate_new1$Compound)
setdiff(Nate_new1$Compound,VOC_new$Compound) # not sure why it isn't removing (E)-beta-ocimene
```
Differences between uafR and Nate for Liz’s data

```{r}
Nate_raw <- mutate_all(as.data.frame(Nate_new1[c(2:60)]), function(x) as.numeric(as.character(x)))
Nate_raw$Compound <- Nate_new1$Compound
Nate_raw$RT <- Nate_new1$RT

VOC_raw <- mutate_all(as.data.frame(VOC_new[c(2:60)]), function(x) as.numeric(as.character(x)))
VOC_raw$Compound <- VOC_new$Compound
VOC_raw$RT <- VOC_new$RT

# Loop through both dataframes for each sample to get differences in retention time and quantity
MAGdiff <- list()
for (i in colnames(Nate_raw)[1:59]){
  sample = i
  df1 <- subset(Nate_raw,select=c("Compound","RT",sample))
  colnames(df1)<-c("Compound","RT","sample")
  df1$sample<-as.numeric(df1$sample)
  df1$data <- "Nate"
  row.names(df1)<-NULL
  if(is.null(VOC_raw[[sample]])){
    next
  }
  df2 <- subset(VOC_raw,select=c("Compound","RT",sample))
  colnames(df2)<-c("Compound","RT","sample")
  df2$sample<-as.numeric(df2$sample)
  df2$data <- "uafR"
  row.names(df2)<-NULL
  df4 <- merge(df1,df2,by="Compound")
  RTdiff <- df4$RT.x - df4$RT.y
  df4$diff2 <- df4$sample.x - df4$sample.y
  MAGdiff[[i]]<-df4$diff2
}

# Make a dataframe showing average differences in RT and quantity
DiffDF <- data.frame(do.call(cbind, MAGdiff))
DiffDF$Compound <- df4$Compound
DiffDF$RTdiff <- RTdiff
DiffDF_melted <- melt(DiffDF, id=c("Compound","RTdiff"))
Differences_df <- DiffDF_melted %>% dplyr::group_by(Compound) %>% dplyr::summarise(RTdiff_avg=mean(RTdiff),MAGdiff_avg=mean(value))

# create a table with compounds that either showed large differences in RT or quantity
Differences_df[(abs(Differences_df$RTdiff_avg)>0.1)|(abs(Differences_df$MAGdiff_avg)>10),] %>% 
  mutate(across(c('RTdiff_avg', 'MAGdiff_avg'), round, 2))%>%
  dplyr::rename("Average difference in retention time" = "RTdiff_avg","Average difference in peak area" = "MAGdiff_avg") %>% 
  gt()

```


In most cases Nate’s proportional quantities are greater than uafR’s, but RT is generally the same
The ones where the quantity is significantly higher in uafR samples but RT is the same:

- 1-dodecanol
- alpha-pinene
-	Methyl salicylate
- n-octane

The ones where the RT and quantity is different (uafR vs Nate)

-	Cumene (8.11 vs 7.5) 0.61
-	Delta-3-carene (9.5  vs 8.99) 0.51
-	Linalool (8.98  vs 10.42) -1.43
- n-tridecan-1-ol (16.6 vs 16.9)
-	P-cymene (10.37 vs. 9.23) 1.14
-	Sabinene (8.75 vs 8.36) 0.39 



